Syntax and Semantics of Cedille

### 2020-12-25 docs/semantics/paper.tex
docs: semantics: minor
- context formation
- types can't have their kinds checked

The classification rules are given in
Figures~\ref{fig:superknd},~\ref{fig:knd}, and~\ref{fig:tp}.  For
brevity, we take these figures as implicitly specifying the syntax of
kinds $\kappa$, types $T$, and annotated terms $t$; these may use term
variables $x$ and type variables $X$, which we assume come from
distinct sets.  So terms and types are syntactically distinguished.
The typing rules (Figure~\ref{fig:tp}) are
bidirectional~\cite{pierce+00}, while the kinding and superkinding
rules (Figure~\ref{fig:knd} and~\ref{fig:superknd}) are only
synthesizing.  We write $\Leftrightarrow$ to range over
$\{\tpcheck,\tpsynth\}$, and when this symbol occurs multiple times in a rule,
it is intended that such occurrences be read the same way (i.e., either all as
\(\tpcheck\) or all as \(\tpsynth\)).
We follow the syntax of our implementation
Cedille, which distinguishes application of a term or type $e$ to a
type ($e \cdot T$), from application to a term ($e\ t$), and
application to an erased term argument ($e\ \mhyph t$, in which case \(e\) must
be a term).
The rules are intended, with a few points of nondeterminism, to be read
bottom-up (in a standard way;
c.f.~\cite{peytonjones07,Pfe01_Lecture-Notes-on-Bidirectional-Type-Checking}) as
an algorithm for synthesizing a classifier from a context and an expression
($\tpsynth$) (\emph{type synthesis}) or checking an expression against a classifier in context
($\tpcheck$) (\emph{type checking}).
Specifically, nondeterminism arises in the two rules allowing both synthesized
and checked types to be reduced, written \(T \leadsto^*_\beta T'\), so that type
constructors required by other rules may be revealed.
The reduction relation \(\leadsto_\beta\) is call-by-name normalization, as this
is sufficient to reduce types to weak head normal form.

@@

\begin{figure}
  \[
  \begin{array}{ll}
    \infer{\Gamma \vdash X \tpsynth \kappa}{(X : \kappa) \in \Gamma} &
    \infer{\Gamma\vdash \abs{\forall}{X}{\kappa}{T} \tpsynth \star}{\Gamma \vdash \kappa & \Gamma,X:\kappa\vdash T \tpsynth \star} \\ \\
    \infer{\Gamma\vdash\abs{\forall}{x}{T}{T'} \tpsynth \star}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T' \tpsynth \star} &
    \infer{\Gamma\vdash\abs{\Pi}{x}{T}{T'} \tpsynth \star}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T' \tpsynth \star} \\ \\
    \infer{\Gamma\vdash\abs{\lambda}{x}{T}{T'} \tpsynth \abs{\Pi}{x}{T}{\kappa}}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T'\tpsynth\kappa} &
    \infer{\Gamma\vdash\abs{\lambda}{X}{\kappa}{T'} \tpsynth \abs{\Pi}{X}{\kappa}{\kappa'}}{\Gamma \vdash \kappa & \Gamma,X:\kappa\vdash T'\tpsynth\kappa'} \\ \\
    \infer{\Gamma\vdash T\ t \tpsynth [t/x]\kappa}{\Gamma\vdash T \tpsynth \abs{\Pi}{x}{T'}{\kappa} & \Gamma\vdash t \tpcheck T'} &
    \infer{\Gamma\vdash T_1 \cdot T_2 \tpsynth [T_2/X]\kappa_1}{\Gamma\vdash T_1 \tpsynth \abs{\Pi}{X}{\kappa_2}{\kappa_1} & \Gamma\vdash T_2 \tpsynth \kappa_2' & \kappa_2 \cong \kappa_2'} \\ \\
    \infer{\Gamma\vdash\abs{\iota}{x}{T}{T'} \tpsynth \star}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T' \tpsynth \star} &
    \infer{\Gamma\vdash \{ t \simeq t' \} : \star}{\textit{FV}(t\ t')\subseteq\textit{dom}(\Gamma)}
  \end{array}
  \]
  \caption{Rules for synthesizing a kind for a type ($\Gamma \vdash T \tpsynth \kappa$)}
  \label{fig:knd}
\end{figure}

@@

    \infer{\Gamma\vdash \absu{\Lambda}{X}{t} \tpcheck \abs{\forall}{X}{\kappa}{T}}{\Gamma,X:\kappa\vdash t \tpcheck T} &
    \infer{\Gamma\vdash t \cdot T' \tpsynth [T'/X]T}
          {\Gamma\vdash t \tpsynth \abs{\forall}{X}{\kappa}{T} & \Gamma\vdash T' \tpcheck\kappa} \\ \\

>>>>

The classification rules are given in
Figures~\ref{fig:superknd},~\ref{fig:knd}, and~\ref{fig:tp}, with
Figure~\ref{fig:ctxt} giving the context formation rules.
For brevity, we take these figures as implicitly specifying the syntax of
contexts \(\Gamma\), kinds $\kappa$, types $T$, and annotated terms $t$; these
may use term variables $x$ and type variables $X$, which we assume come from
distinct sets.  So terms and types are syntactically distinguished.
The typing rules (Figure~\ref{fig:tp}) are
bidirectional~\cite{pierce+00}, while the kinding and superkinding
rules (Figure~\ref{fig:knd} and~\ref{fig:superknd}) are only
synthesizing.  We write $\Leftrightarrow$ to range over
$\{\tpcheck,\tpsynth\}$, and when this symbol occurs multiple times in a rule,
it is intended that such occurrences be read the same way (i.e., either all as
\(\tpcheck\) or all as \(\tpsynth\)).
We follow the syntax of our implementation
Cedille, which distinguishes application of a term or type $e$ to a
type ($e \cdot T$), from application to a term ($e\ t$), and
application to an erased term argument ($e\ \mhyph t$, in which case \(e\) must
be a term).
The rules are intended, with a few points of nondeterminism, to be read
bottom-up (in a standard way;
c.f.~\cite{peytonjones07,Pfe01_Lecture-Notes-on-Bidirectional-Type-Checking}) as
an algorithm for synthesizing a classifier from a context and an expression
($\tpsynth$) (\emph{type synthesis}) or checking an expression against a classifier in context
($\tpcheck$) (\emph{type checking}).
Specifically, nondeterminism arises in the two rules allowing both synthesized
and checked types to be reduced, written \(T \leadsto^*_\beta T'\), so that type
constructors required by other rules may be revealed.
The reduction relation \(\leadsto_\beta\) is call-by-name normalization, as this
is sufficient to reduce types to weak head normal form.

@@

\begin{figure}
  \[
  \begin{array}{ll}
    \infer{\Gamma \vdash X \tpsynth \kappa}{(X : \kappa) \in \Gamma} &
    \infer{\Gamma\vdash \abs{\forall}{X}{\kappa}{T} \tpsynth \star}{\Gamma \vdash \kappa & \Gamma,X:\kappa\vdash T \tpsynth \star} \\ \\
    \infer{\Gamma\vdash\abs{\forall}{x}{T}{T'} \tpsynth \star}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T' \tpsynth \star} &
    \infer{\Gamma\vdash\abs{\Pi}{x}{T}{T'} \tpsynth \star}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T' \tpsynth \star} \\ \\
    \infer{\Gamma\vdash\abs{\lambda}{x}{T}{T'} \tpsynth \abs{\Pi}{x}{T}{\kappa}}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T'\tpsynth\kappa} &
    \infer{\Gamma\vdash\abs{\lambda}{X}{\kappa}{T'} \tpsynth \abs{\Pi}{X}{\kappa}{\kappa'}}{\Gamma \vdash \kappa & \Gamma,X:\kappa\vdash T'\tpsynth\kappa'} \\ \\
    \infer{\Gamma\vdash T\ t \tpsynth [t/x]\kappa}{\Gamma\vdash T \tpsynth \abs{\Pi}{x}{T'}{\kappa} & \Gamma\vdash t \tpcheck T'} &
    \infer{\Gamma\vdash T_1 \cdot T_2 \tpsynth [T_2/X]\kappa_1}{\Gamma\vdash T_1 \tpsynth \abs{\Pi}{X}{\kappa_2}{\kappa_1} & \Gamma\vdash T_2 \tpsynth \kappa_2' & \kappa_2 \cong \kappa_2'} \\ \\
    \infer{\Gamma\vdash\abs{\iota}{x}{T}{T'} \tpsynth \star}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T' \tpsynth \star} &
    \infer{\Gamma\vdash \{ t \simeq t' \} \tpsynth \star}{\textit{FV}(t\ t')\subseteq\textit{dom}(\Gamma)}
  \end{array}
  \]
  \caption{Rules for synthesizing a kind for a type ($\Gamma \vdash T \tpsynth \kappa$)}
  \label{fig:knd}
\end{figure}

@@

    \infer{\Gamma\vdash \absu{\Lambda}{X}{t} \tpcheck \abs{\forall}{X}{\kappa}{T}}{\Gamma,X:\kappa\vdash t \tpcheck T} &
    \infer{\Gamma\vdash t \cdot T' \tpsynth [T'/X]T}
          {\Gamma\vdash t \tpsynth \abs{\forall}{X}{\kappa}{T} & \Gamma\vdash T' \tpsynth \kappa' & \kappa'\cong\kappa} \\ \\
